---
title: "Comparing total lengths of female *Cnemidophorus uniparens* and *Cnemidophorus tessalatus* "
author: "Cristina Mancilla"
date: "11/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
#install necessary packages

library(tidyverse)
library(here)
library(janitor)
library(broom)
library(dplyr)
library(kableExtra)
library(equatiomatic)

```

### Introduction 

The Cnemidophorus uniparens, also known as the Desert Grassland lizard and Cnemidophorus tessalatus, or the Colorado Checkered lizard, can both be found in the southwestern United States. I will compare the total lengths (from snout to tail) of the two species from data collected near the San Andres National Wildlife Refuge by David Lightfoot in 2017. I will questions: "Is there a significant difference between total lenghts between C.tessalatus and C. uniparens?" using an unpaired, two-sample t-test. Only data for female lizards will be included because C.uniparens is an all-female species.

```{r}
lizard <- read_csv(here("data", "lizards.csv"))
```


```{r}
# select columns sex, total_length, and spp (species) and filter for observation of only CNUN (Cnemidophorus uniparens, the Desert Grassland lizard) and CNTE (Cnemidophorus tessalatus, the Colorado Checkered lizard).


lizard_fix <- lizard %>% 
  select("sex", "total_length", "spp") %>% 
  filter(spp %in% c("CNUN", "CNTE")) %>% 
  filter(sex =="F") %>% 
  drop_na()

  
```



```{r, eval = FALSE}
mu <- ddply(lizard_fix, "spp", summarise, grp.mean=mean(total_length))
head(mu)
```
### Analysis 

```{r}
# create a visualization comparing total length for female CNTE and CNUN lizards 

ggplot(data = lizard_fix, aes(x= total_length, color = spp)) +
  geom_histogram(bins= 20,color = "white") +
  aes(color = spp, fill = spp) +
  
  
# add labels 
  
  labs( x = "Total length (mm)", 
        y = "counts",
        title = "Total length (mm) of female Cnemidophorus uniparens (CNUN)\n and Cnemidophorus tessalatus (CNTE) lizards", 
        caption = "Comparison of the total length (mm), from snout to end of tail, of female Cnemidophorous uniparens \n (Desert Grassland Lizard) and Cnemidophorus tessalatus (Colorado Checkered Lizard).Dashed lines \n represent mean total length (mm) for each species of lizard.The mean for C.uniparens (CNUN) is 147 mm \n and C.tessalatus (CNTE) is 245 (mm) Lizards were sampled using pitfall traps near the San Andres National \n Wildlife Refuge.") +
  
#customize title 
  
theme(plot.title = element_text(color = "midnightblue", size = 14, face = "bold"),  
      plot.caption = element_text(hjust = 0, size = 10)
  ) +
  
geom_vline(xintercept = 244.89, linetype="dotted", 
                color = "coral", size=1.0) +
  
geom_vline(xintercept = 147, linetype="dotted", yend= 0, 
                color = "turquoise4", size=1.0)
  
  

         
```
```{r}
lizards_summarize <-lizard_fix %>% 
  
  group_by(spp)%>% #group by species
  
  summarize(mean_mm = round(mean(total_length), 1), 
          sd_mm = round(sd(total_length), 1), 
          median_mm = round(median(total_length), 1), 
          n = n()) %>% 
  
#use Kble to add a table caption and update the theme and style of the table so that it looks presentable when it is knitted. 
  
  kbl(caption ="Summary statistics for total length of female C.uniparens (CNUN) and C. tessalatus (CNTE) " ) %>% 
  
kable_classic(full_width = F, html_font = "Cambria") #this is the theme of the table 

  lizards_summarize
 #call the object back to view the table 
```



\n 
\n 



The mean length of C.uniparens (CNUN) is 244.9 (mm) ± 47.3 and the mean total length for C. tessalatus (CNTE) is 147.6 (mm) ± 34.6. It's important to note that there are more observations for CNUN (40) than CNTE (28). 

 
```{r}
#create subsets to use for t-tests and exploratory plots

cnte <- lizard_fix %>% 
  filter(spp=="CNTE") %>% 
  pull(total_length) 


cnun <- lizard_fix %>% 
  filter(spp=="CNUN") %>% 
  pull(total_length)
```


```{r, include = FALSE}
#visualize the data using histograms to check for normality - appropriate for t-test? Both subsets look normally distributed, so the t-test assumption of normal distribution is accurate and a t-test is appropriate. 

hist(cnte)
hist(cnun)
```


```{r, include = FALSE}
# t-test! 

lizard_ttest <- t.test(cnte,cnun,na.rm=T)
```

```{r, include = FALSE}
# cohen's D  for effect size 

lizard_cohen <- effsize::cohen.d(cnte, cnun,na.rm=T)
```

There is a significant difference between total lengths (mm) of female C.uniparens (CNUN) and C. tessalatus (CNTE) (Welch's two-sample, unpaired t-test: t(`r lizard_ttest$statistic`), p = `r lizard_ttest$p.value`) (p<0.05) . The absolute difference in means between the two species is 97.3 (mm).The effect size is large with a difference of (Cohen's d = `r lizard_cohen$estimate`) standard deviations between the species total lengths. Data is normally distributed for both groups.  

         
         
         
Data source: Lightfoot, D. 2017. Lizard pitfall trap data (LTER-II, LTER-III) ver 36. Environmental Data Initiative. https://doi.org/10.6073/pasta/ff37b4cace16a9943575f3fd7067064e



