---
title: "Exploring the association between political affiliation and self-identification as an 'Environmentalist' "
author: "Cristina Mancilla"
date: "11/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,warning = FALSE )

#attach necessary packages

library(here)
library(tidyverse)
library(janitor)
library(broom)
library(dplyr)
library(kableExtra)

```
### Introduction: 

The Andrew Forest Long-Term Ecological Research site published a report led by Chelsea Batavia that recorded survey data for an online panel of 1,331 anonymous, nationwide U.S respondents.Data collection was completed in 2017 and formally published in 2019. I will use information recorded about survey respondents' political affiliation and self-identification as an "Environmentalist" to determine if there is any association between the two identifiers. I will use a Chi-square test (X^2^) to find evidence of association. 




```{r}
# read in the data! 

conservation <- read_csv(here("data", "conservation_survey.csv"))

#Filter the data so that only variables ENVI_ID and POL_OR are included in the dataframe. 

```


```{r}
#Add a column that bins political party into a binary response of either "conservative" (1-3) or "liberal" (5-7). Exclude all other political affiliations. 

envi_pol <- conservation %>% 
  select("ENV_ID", "POL_OR") 
  
```


```{r}
envi_pol_out <- envi_pol %>% 
  mutate(POL_OR = case_when(
    POL_OR %in% c(1, 2, 3) ~ "Conservative", 
    POL_OR %in% c(5, 6, 7) ~ "Liberal"
    
  )) %>% 
  
  mutate(ENV_ID = case_when(
    ENV_ID %in% c(1) ~ "Yes", 
    ENV_ID %in% c(2) ~ "No", 
    ENV_ID %in% c(3) ~ "Unsure"
  )) 

```


```{r, include = FALSE}
#count the occurrences of political affilication and environmentalist identification 

envi_pol_out %>% 
  count(ENV_ID, POL_OR)
```

```{r}
envipol_counts <- envi_pol_out %>% 
  na.omit(enviipol_counts) %>% # remove NA from contingency table counts 
  janitor::tabyl(ENV_ID, POL_OR)

# make another version of this table that has the counts AND proportions. 

envipol_proportions <- envipol_counts %>% 
  janitor::adorn_percentages() %>% 
  janitor::adorn_pct_formatting(digits = 2) %>% # put it in a percentage format
  janitor::adorn_ns()  # add actual counts in each cell 

kable(envipol_proportions, 
  caption =" Association between respondents' political affiliation and self-identification as an 'environmentalist' ") %>% 
  kable_styling(font_size = 15) %>%
  kable_material(full_width = FALSE, html_font = "Cambria") # update the table theme
  

```
Counts and proportions of respondents who identified as either 'liberal' or 'conservative' and who responded “Yes”, “No” and “Unsure” to the statement “I consider myself an environmentalist.”

```{r}

# convert the contingency table into a format that can be correctly analyzed in a chi-square test. Use column_to_rownames() to move the ENV_ID column into a row name. 

envi_pol_ct_final <- envipol_counts %>% 
  column_to_rownames(var = "ENV_ID")
```

### Analysis: Chi-Square test 


```{r, include = FALSE}
envipo_x2 <- chisq.test(envi_pol_ct_final)
envipo_x2

envipo_tidy <- broom ::tidy(envipo_x2)
```
Using Pearson's Chi-square test to test for significance, I found a significant association (p<0.05) between political affiliation and self - identification as an environmentalist ($Chi$^2^( `r envipo_tidy$parameter` ) = `r envipo_tidy$statistic`, *p* = `r envipo_tidy$p.value `). Many respondents who identified as "conservative" did not identify as an "environmentalist" (71.3%). In comparison, only 28% of "liberals" also responded "No". Although the difference is not as drastic, more "liberals" responded "Yes" (56.88%) to identifying as an "environmentalist" than "conservatives" (43.12%).   


Data citation: Andrews Forest LTER Site and C. Batavia. 2019. Influences on charitable giving for conservation: Online survey data of 1,331 respondents across the US, August 2017 ver 3. Environmental Data Initiative. https://doi.org/10.6073/pasta/926e6270e324a1322a900da14d38b96c
